Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-11 Multiresolution_framework/ltx/manuscript.tex
moved M-step to the appendix

@@

The rest of this section is dedicated to the construction of the lower bound, $\mathcal Q$, for the linear state-space model given by Eqs.~\eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}. The joint probability distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta)$ can be written as
\boldsymbol\Xi_0&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \label{eq:MRA-Xi0}\\
\boldsymbol\Xi_1&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right] \label{eq:MRA-Xi1}.
The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \citep{Gibsona2005}, (see \ref{ap:Xiderivation} for more details)
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:Xi0} \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).  \label{eq:Xi1}
\begin{equation}\label{eq:MRA-QCompact}
\begin{equation}\label{eq:epsilon0}
\begin{equation}\label{eq:epsilon1}
\begin{equation}\label{eq:Epsilon}
The second derivative of the $\mathcal Q$-function is given by
\frac{\partial^2\mathcal Q}{\partial\boldsymbol\theta^2}=-2T_s\varsigma\boldsymbol\Upsilon.
The matrix $\boldsymbol\Upsilon$ is positive definite if $n_x^2\times n_{\theta}$ matrix $\mathrm{vec}(\mathbf U)$ is of rank $n_{\theta}$ (\citep{Dewar2009}, Lemma 3), where $\mathrm{vec}(\cdot)$ denotes the vectorization operator. Under this condition $\boldsymbol\Upsilon$ is  invertible and the second derivative is negative definite representing a maximum of the $\mathcal Q$-function.

@@

The EM algorithm alternates between maximizing the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximized with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximizing the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in Eq.~\eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from Eq.~\eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximizes the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in Eq.~\eqref{eq:MRA-Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by computing the cross-covariance matrix as

>>>>

\begin{equation}\label{eq:Bishopbound}

@@

For the linear state-space model given by Eqs.~\eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound defined by Eq.~\eqref{eq:Bishopbound} is a quadratic of the form (see \ref{ap:QDerivation} for derivation) 
\begin{equation}\label{eq:MRA-QCompact}
\mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\varsigma\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\varsigma\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
where $\beta$ is constant with respect to $\boldsymbol\theta$. Equation~\eqref{eq:MRA-QCompact} is maximum at
\begin{align}\label{eq:MRA-thetahat}
\boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1)^\top,
\end{align}
where
\begin{align}
\boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\boldsymbol\Gamma_1]_{j,i}\label{eq:epsilon0} \\
\boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Gamma_1]^{j,i} \label{eq:epsilon1}\\
\boldsymbol\Upsilon&=T_s\varsigma\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Gamma_2]^{j,i},\label{eq:Epsilon}
\end{align} 
where $[\cdot]_{p,q}$ denotes the $\left(p,q\right)$-element of the matrix, and $ [\cdot]^{p,q}$ denotes the $\left(p,q\right)$-block of the block matrix and where 
\begin{align}
\left[ \Gamma_1\right]^{j,i} &=\sum_{k=1}^{n_x}\left[ \boldsymbol\Sigma_w^{-1}\boldsymbol\Lambda_x^{-1}\right]_{jk} \left[ \mathbf U\right]^{k,i},\\
\left[ \Gamma_2\right] ^{j,i}&=\sum_{k,m=1}^{n_x}[\mathbf U^{\top}]^{jk} \left[\boldsymbol\Lambda_x^{-1}\boldsymbol\Sigma_w^{-1}\boldsymbol\Lambda_x^{-1} \right]_{km}[\mathbf U]^{mi}.
\end{align}
Each $ 1 \times n_{\theta}$ block of the $n_x \times n_x n_{\theta}$ block matrix $\mathbf U$ is 
\begin{align}\label{eq:Uij}
\left[ \mathbf U\right] ^{k,i}&=\int_{\boldsymbol \Omega}\left[\boldsymbol\mu(r) \right]_k \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(r-r'\right) \mathrm{d}r'\right]_{i:} \mathrm{d}r,
\end{align}
where $[\cdot]_{i:} $ denotes the $i^{th}$ row.  Each element of $\boldsymbol\Lambda_{\theta}$ can be calculated using
\begin{equation}\label{eq:MRA-LambdaThetaElements}
\left[ \boldsymbol\Lambda_{\theta}\right] _{k,i}=\left[ \mathbf U\right]^{k,i}\boldsymbol\theta.
\end{equation}
The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ in Eqs.~\eqref{eq:epsilon0}-\eqref{eq:Epsilon} are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \citep{Gibsona2005}, 
\begin{align}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:Xi0} \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).  \label{eq:Xi1}
\end{align} 
The second derivative of the $\mathcal Q$-function is given by
\begin{equation}
\frac{\partial^2\mathcal Q}{\partial\boldsymbol\theta^2}=-2T_s\varsigma\boldsymbol\Upsilon.
\end{equation}
The matrix $\boldsymbol\Upsilon$ is positive definite if $n_x^2\times n_{\theta}$ matrix $\mathrm{vec}(\mathbf U)$ is of rank $n_{\theta}$ (\citep{Dewar2009}, Lemma 3), where $\mathrm{vec}(\cdot)$ denotes the vectorization operator. Under this condition $\boldsymbol\Upsilon$ is  invertible and the second derivative is negative definite representing a maximum of the $\mathcal Q$-function.

@@

The EM algorithm alternates between maximizing the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximized with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximizing the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in Eq.~\eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from Eq.~\eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximizes the expectation of the complete data log-likelihood.

\section{Lower bound}\label{ap:QDerivation}
For the linear state-space model given by Eqs.~\eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}. The joint probability distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta)$ can be written as
\boldsymbol\Xi_0&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \label{eq:app-Xi0}\\
\boldsymbol\Xi_1&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right] \label{eq:app-Xi1}.
The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ are calculated using the RTS smoother outputs (see \ref{ap:Xiderivation} for more details)
\begin{equation}\label{eq:appQCompact}
\begin{equation}\label{eq:appepsilon0}
\begin{equation}\label{eq:appepsilon1}
\begin{equation}\label{eq:appEpsilon}
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in Eq.~\eqref{eq:MRA-Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	We start by computing the cross-covariance matrix as
